\documentclass[11pt]{article}
\usepackage{framed}

\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{tikz}
\usepackage{amsmath}
\lstset{language=C}
\begin{document}

\title{Survivability of ICU Patients with Severe Sepsis/Septic Shock}
\author{Jeremy B. Crowe - crowe.jb@gmail.com}
\maketitle

\section{Domain Background}
\subsection{Overview}
Sepsis is a serious, life threatening condition that is caused by an overwhelming immune response to an infection. Infection is often bacterial however it is possible for sepsis to be associated with: fungal, viral, parasitic, as well as bacterial infections.

Sepsis is due to the body's response, not only the effects of the infection itself. White blood cells release an array of chemicals to fight the infection which trigger systemic inflammation, vasodilation, permeability of vessels, and intracellular fluid build up. These ``leaky vessels'' deplete the body of coagulation factors. The increased fluid build up and decreased blood pressure result in a lack of oxygenation of tissue, known as shock.

If sepsis is not treated quickly or with enough direct care, multiple organ dysfunction can occur which can result in kidney failure, liver failure, heart failure, acute respiratory distress, etc. The speed at which treatment is administered and types of treatments are directly correlated with survivability of this severe condition. The speed of treatment has been found to be more important than the age of the patient~\cite{survival2}. Each hour of delay in antimicrobial administration over the initial 6 hours is associated with an average decrease in survival rate of 7.6\%~\cite{survival}.

Certain treatments may be highly effective, therefore it is important to pay special attention to several things: the type of infection and type of antibiotics used, the blood pressure and whether or not vasopressors were used, correct amount of IV fluid provided to patient. The correct type of antibiotic and appropriate administering of vasopressors can reduce the chance of organ failure and mortality greatly~\cite{pressors}.

\subsection{Why This is Important}

Sepsis is partially responsible for 6\% of all deaths in the United States~\cite{cdc}. Sepsis is a highly variable condition caused by numerous infections and pre-existing conditions. A patient may arrive at the ICU with sepsis or may be diagnosed during the stay. There is no diagnostic test to confirm or deny the existence of such a condition in a patient. It is diagnosed based on the clinical judgment of physician. Regardless of the cause, the speed at which treatment is administered and the types of treatments are critical in stabilizing a patients blood pressure, body temperature, eliminate infection, and keep cardiac output stable. Due to the variable, and difficult nature of the condition a machine learning approach may be able to aid in predicting the severity of the condition and the effectiveness of potential treatments.


\section{Problem}

A predictive model would be highly useful for ICU staff and physicians. It would aid in prioritizing patients that are considered ``high risk'' and as a simple way to stratify patients in clinical research. 

Upon patient admittance and initial blood and microbiology tests patient survivability will be classified over a 30 day period. While survival rates are available long after 30 days, this is the most critical period for surviving a condition such as sepsis/septic shock. 

Several class features will be considered: gender, ethnicity, culture classification, antibiotic treatment, etc. Several continuous features will also be considered: time before first antibiotic dose, age, respiratory rate, blood lactate levels, creatinine levels, blood pressure, etc.

The survivability of sepsis is strongly correlated with the speed of treatment, age of patient, type of infection, and general state of the condition. The state of the condition can be measured in a variety of ways. Respiratory rate, blood pressure, creatinine (kidneys failing), and lactate levels(anaerobic cellular respiration) strongly suggest the current stage of the condition. Upon arrival these measurements can be used to classify a patient into a survival or not survival group.

\section{Datasets and Inputs}
For this study I will be using the MIMIC-III dataset(https://mimic.physionet.org/). This dataset is freely accessible and suggested for use by Udacity. I will be utilizing the data on more than 40,000 individual patients. 1184 patients were admitted with sepsis and thousands more diagnosed by the end of their stay. I will be focusing on the patients admitted with sepsis since I have a time of admittance that will aid in finding more effective treatments. Diagnosis data is tied to ICD-9 billing codes that do not have a time associated.

It was required to take a short course to gain access to this data by demonstrating some understanding of law and ethics in dealing with this de-identified dataset.

I have downloaded the data locally as csv files. I then modified a script provided by MIT to load the entries into an Amazon RDS Postgres database. This database and server is secured and is HIPAA compliant. I then have spun up an Amazon compute EC2 instance to do most of the heavy lifting. I will be using many tables from the database but the most critical are the following:
\begin{enumerate}
\item \textbf{admissions}: the admissions table provides basic patient information on admission including the initial diagnosis, sex, insurance type, etc.
\item \textbf{diagnoses\_icd}: this table provides all diagnoses assigned to a patient during a hospital stay. Unfortunately this is assigned at the end and cannot be given a time of diagnosis. A sequence is provided however.
\item \textbf{microbiologyevents}: this table provides information about whether or not an infection is present, how it was obtained, and when.
\item \textbf{labevents}: this provides measurements such as blood pressure, and information from blood tests.
\item \textbf{prescriptions}: drugs given to patient along with the time and amount.
\item \textbf{patients}: this provides basic information about the patient including date of birth, date of death (if applicable), gender, etc.
\end{enumerate}

In this supervised learning problem, the target attribute must be calculated from the provided information. The goal is to predict survivability within 30 days of admittance. Within the patient table exists a date of death (dod), if applicable, that can be used along with admittance date to calculate the boolean survivability target attribute.


\section{Solution}
The solution to is to categorizing patients by survivability at admittance. This is calculated by considering: speed of first antibiotic dose, age of patient, type of infection, respiratory rate, blood pressure, creatinine levels, and lactate levels. A supervised learning model will be trained off of this information along with the calculated target value, which is the patient survival over 30 days from admittance.

As the project progresses it may require me to include more complex information such as blood pressure measurements and categorical data about whether vasopressors were used to combat low bp and more.
Multiple models using different techniques will be developed, analyzed and a comparative analysis between them will be done with the best performing and most generalizable solution being selected.

\section{Benchmark and Evaluation}
There are few studies into the predictability of patient survival from severe sepsis. There are also many factors that are important in how the prediction is made. I will be focusing on the following research: ``Predicting survival of patients with sepsis by use of regression and neural network models'' by J.R. Flanagan, et al. \begin{quotation}
  ``The development of multiple organ failure after the diagnosis of sepsis was predicted accurately (81.5\%) with either the MLR or the NN model. Both the MLR and the NN methods depended on the interpretation of a likelihood quantity, requiring the choice of a threshold to make a survival prediction.''\cite{sepsisresearch}
\end{quotation}

While this study was done in 1996, there have been few followup studies. The few that I found had a similar success rate. Due to the detail of this study I am focusing on the work of Flanagan et al. for my benchmark. I have requested full access to their research via ResearchGateway.

\section{Design}
Due to the nature of the problem and the relatively large number of features that will likely be involved, I will be comparing the results of a Neural Network, Adaboost with a decision stump as its base estimator or extreme gradient boosting. Only Adaboost is possible in SKlearn, so I may need special permission to use Tensorflow for a multi-layer Neural Network and XGBoost for extreme gradient boosting.

The data is spread across multiple database tables. For most algorithms I will need to join this data into a large single entry if not using time series data. For reusability I will likely create a new table with this joined information for easy querying.


\bibliography{citations}{}
\bibliographystyle{ieeetr}

\end{document}